---
title: "Hurricane Lane: Power Outages Seen Through Nighttime Lights"
subtitle: "Hawaii, August–September 2018 | NASA Black Marble VIIRS"
author: "Izack Takazawa"
format:
  html:
    toc: true
format: html
editor: visual
---

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(tigris)
library(sf)
library(blackmarbler)
library(terra)
library(tidyterra)
library(lubridate)
library(exactextractr)
library(here)
library(gt)
library(scales)
#install.packages("patchwork")
library(patchwork)

options(tigris_use_cache = TRUE)
options(scipen = 999)
```

# Research question

**How did nighttime radiance across Hawaii change during Hurricane Lane?**

Hurricane Lane at its peak was a Category 5 hurricane. Although technically Lane never made landfall, its proximity to the islands caused torrential rainfall.

Big Island received the most notable rainfall averaging around 17 inches of rain over four days according to NOAA.

This analysis compares a pre-storm nighttime-light baseline with the Hurricane Lane period and week after which I labeled "restoration period".

To best give a proper analysis upon the storms effect of power outages we chose to analyze both before, during, and after the storm.

# Setting up Dates/Timeframe

```{r TIMEFRAME}
# Timeframe of analysis
#Pre-storm (1 week prior) August 1-8 2018
#During the Storm August 15-29 
#Post storm ("Restoration Period") August 30-September 15)
baseline_dates <- seq.Date(as.Date("2018-08-01"), as.Date("2018-08-08"), by = "day")
storm_dates <- seq.Date(as.Date("2018-08-15"), as.Date("2018-08-29"), by = "day")
restoration_start <- as.Date("2018-08-30")
analysis_end <- as.Date("2018-09-15")
restoration_dates <- seq.Date(restoration_start, analysis_end, by = "day")
all_dates <- c(baseline_dates, storm_dates, restoration_dates)

#TOKEN
bearer <- "eyJ0eXAiOiJKV1QiLCJvcmlnaW4iOiJFYXJ0aGRhdGEgTG9naW4iLCJzaWciOiJlZGxqd3RwdWJrZXlfb3BzIiwiYWxnIjoiUlMyNTYifQ.eyJ0eXBlIjoiVXNlciIsInVpZCI6Iml6YWNrdGFrYXphd2EiLCJleHAiOjE3OTI4Nzg0OTQsImlhdCI6MTc4NzY5NDQ5NCwiaXNzIjoiaHR0cHM6Ly91cnMuZWFydGhkYXRhLm5hc2EuZ292IiwiaWRlbnRpdHlfcHJvdmlkZXIiOiJlZGxfb3BzIiwiYWNyIjoiZWRsIiwiYXNzdXJhbmNlX2xldmVsIjozfQ.K2gvioISpYOeE3LyJWf0yE7s4Bo0UedQd-SSqR0MWvteifBbC3l18ptdSm14OC12D-k2r7aapajadzF20W3zIS3WWrjn8gtcYD6wXKDKJSwRvf-9C0gM5Gg4NZ1LYCitMTMWtFNnpLiq8pkagDw_z9osbbzyfj_ta2BKTPwIOq8g2ktKTIzHLWTuNagdRFptZRl3cZRo-aPCY_zeW6CAZ5VmjjtEQ9EC9mCHm4tgMEn6HBAKIA6smh6_l4aF3HzhB1qjhWFjyNL6cyFF5zAgdPRz9GgWuBNM9EsACCG_C2QBPqfBwznzl_cJJGDzNW00dizMSBFzOZJrsaHtx_gLhA"

period_lookup <- tibble(
  date = all_dates,
  period = case_when(
    date %in% baseline_dates ~ "Baseline",
    date %in% storm_dates ~ "Storm period",
    date %in% restoration_dates ~ "Restoration"
  )
) |>
  mutate(period = factor(period, levels = c("Baseline", "Storm period", "Restoration")))
```

# Where we are looking to analyze

Hawaii County

Maui County

Honolulu County (Oahu)

Kauai County

```{r counties-map}
roi_hawaii <- counties(
  state = "HI",
  cb = TRUE,
  year = 2018
) |>
  st_transform(4326) |>
  st_crop(
    xmin = -160.5,
    xmax = -154.5,
    ymin = 18.5,
    ymax = 22.5
  ) |>
  filter(NAME != "Kalawao") |>
  mutate(
    county_label = case_when(
      NAME == "Hawaii" ~ "Hawaii County",
      NAME == "Maui" ~ "Maui County",
      NAME == "Honolulu" ~ "Honolulu County (Oahu)",
      NAME == "Kauai" ~ "Kauai County",
      TRUE ~ NAME
    )
  )

ggplot(roi_hawaii) +
  geom_sf(
    aes(fill = county_label),
    color = "white",
    linewidth = 0.5
  ) +
  geom_sf_text(
    aes(label = county_label),
    size = 3,
    color = "black"
  ) +
  scale_fill_brewer(
    palette = "Set2",
    name = "County"
  ) +
  coord_sf(
    xlim = c(-160.5, -154.5),
    ylim = c(18.5, 22.5),
    expand = FALSE
  ) +
  labs(
    title = "Hawaii Counties Included in the Analysis",
    subtitle = "Hawaii County, Maui County, Honolulu County (Oahu), and Kauai County",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

## Specifically on Oahu we will be looking at the Oahu sub areas:

Honolulu

Windward Oahu

Leeward Oahu

North Shore

# Pull Black Marble data

```{r pull-rasters}
pull_one_date <- function(d) {
  tryCatch(
    bm_raster(
      roi_sf = roi_hawaii,
      product_id = "VNP46A2",
      date = d,
      bearer = bearer,
      quality_flag_rm = 2
    ),
    error = function(e) {
      message("Failed for ", d)
      NULL
    }
  )
}

r_list <- purrr::map(all_dates, pull_one_date)

names(r_list) <- as.character(all_dates)

r_list <- purrr::compact(r_list)

message("Downloaded dates: ", paste(names(r_list), collapse = ", "))

message(
  "Dropped dates: ",
  paste(
    setdiff(as.character(all_dates), names(r_list)),
    collapse = ", "
  )
)

r_all <- rast(r_list) |>
  mask(roi_hawaii)

names(r_all) <- names(r_list)

available_dates <- as.Date(names(r_all))

available_lookup <- period_lookup |>
  filter(date %in% available_dates)
```

# County/island radiance

Display all of Hawaii counties included in the analysis

```{r county-radiance}
extract_daily_means <- function(raster_stack, polygons, area_name) {
  purrr::map_dfr(seq_len(nlyr(raster_stack)), function(i) {
    tibble(
      area = polygons[[area_name]],
      date = as.Date(names(raster_stack)[i]),
      radiance = exact_extract(raster_stack[[i]], polygons, fun = "mean")
    )
  })
}

county_daily <- extract_daily_means(r_all, roi_hawaii, "island") |>
  left_join(period_lookup, by = "date")

county_summary <- county_daily |>
  filter(period %in% c("Baseline", "Storm period")) |>
  group_by(area, period) |>
  summarize(
    mean_radiance = mean(radiance, na.rm = TRUE),
    n_nights = sum(!is.na(radiance)),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = period, values_from = mean_radiance) |>
  mutate(
    absolute_change = `Storm period` - Baseline,
    percent_change = 100 * absolute_change / Baseline
  )

county_summary |>
  gt() |>
  fmt_number(columns = c(Baseline, `Storm period`, absolute_change), decimals = 2) |>
  fmt_number(columns = percent_change, decimals = 1, suffixing = FALSE) |>
  cols_label(
    area = "County/island",
    `Storm period` = "Storm-period mean",
    percent_change = "Percent change"
  ) |>
  tab_header(title = "County/Island Baseline Versus Storm-Period Radiance")
```

## 

```{r county-bars}
ggplot(
  county_daily |> filter(period %in% c("Baseline", "Storm period")) |>
    group_by(area, period) |>
    summarize(mean_radiance = mean(radiance, na.rm = TRUE), .groups = "drop"),
  aes(x = area, y = mean_radiance, fill = period)
) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  scale_fill_manual(values = c("Baseline" = "grey65", "Storm period" = "firebrick")) +
  labs(
    title = "Mean Nighttime Radiance by County/Island",
    subtitle = "Pre-storm baseline (Aug 1–8) compared with Hurricane Lane period (Aug 15–29)",
    x = NULL, y = "Mean radiance", fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r county-radiance-timeline}
ggplot(
  county_daily,
  aes(
    x = date,
    y = radiance,
    color = area,
    group = area
  )
) +
  geom_rect(
    data = tibble(
      xmin = min(storm_dates),
      xmax = max(storm_dates),
      ymin = -Inf,
      ymax = Inf
    ),
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = ymin,
      ymax = ymax
    ),
    inherit.aes = FALSE,
    fill = "firebrick",
    alpha = 0.10
  ) +
  geom_line(
    linewidth = 0.7
  ) +
  geom_point(
    size = 1.8
  ) +
  geom_vline(
    xintercept = restoration_start,
    linetype = "dashed",
    color = "steelblue4",
    linewidth = 0.7
  ) +
  scale_x_date(
    date_breaks = "3 days",
    date_labels = "%b %d"
  ) +
  labs(
    title = "Nighttime Radiance by County/Island Through Hurricane Lane",
    subtitle = "Shaded: Hurricane Lane Impact; Line: restoration period",
    x = NULL,
    y = "Mean nighttime radiance",
    color = "County/island"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(
      angle = 35,
      hjust = 1
    )
  )
```

```{r statewide-nighttime-lights-map}
# Choose one successfully downloaded Hurricane Lane date.
map_date <- "2018-08-22"
            #"2018-08-23"
            #"2018-08-24"
            #"2018-08-25"
            #"2018-08-26"

# Check that the requested date exists in your raster stack.
if (!map_date %in% names(r_all)) {
  stop(
    paste0(
      map_date,
      " was not downloaded. Choose a date from: ",
      paste(names(r_all), collapse = ", ")
    )
  )
}

# Pull one daily VNP46A2 raster from your existing stack.
r_map <- r_all[[map_date]] |>
  terra::mask(roi_hawaii)

# Log-transform only the copy used for this map.
# Adding 1 prevents log(0).
r_map[] <- log(r_map[] + 1)

ggplot() +
  geom_spatraster(data = r_map) +
  geom_sf(
    data = roi_hawaii,
    fill = NA,
    color = "white",
    linewidth = 0.35
  ) +
  scale_fill_gradient2(
    low = "black",
    mid = "yellow",
    high = "red",
    midpoint = 4.5,
    na.value = "transparent",
    name = "Log radiance"
  ) +
  coord_sf(
    xlim = c(-160.5, -154.5),
    ylim = c(18.5, 22.5),
    expand = FALSE
  ) +
  labs(
    title = "Hawaii Nighttime Lights During Hurricane Lane",
    subtitle = paste0(
      "NASA Black Marble VNP46A2 daily radiance — ",
      format(as.Date(map_date), "%B %d, %Y")
    )
  ) +
  theme_void() +
  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      hjust = 0.5
    ),
    legend.position = "bottom"
  )
```

```{r county-percent-change}
county_ranked <- county_summary |>
  arrange(percent_change) |>
  mutate(
    area = factor(area, levels = area)
  )

ggplot(
  county_ranked,
  aes(
    x = area,
    y = percent_change,
    fill = percent_change < 0
  )
) +
  geom_col(width = 0.68) +
  geom_hline(
    yintercept = 0,
    color = "grey35",
    linewidth = 0.4
  ) +
  geom_text(
    aes(
      label = paste0(round(percent_change, 1), "%")
    ),
    vjust = ifelse(county_ranked$percent_change < 0, 1.3, -0.35),
    size = 3.6
  ) +
  scale_fill_manual(
    values = c(
      "TRUE" = "firebrick",
      "FALSE" = "darkgreen"
    ),
    guide = "none"
  ) +
  coord_flip() +
  labs(
    title = "County/Island Change in Nighttime Radiance",
    subtitle = "Percent change from the pre-storm baseline to the Hurricane Lane period",
    x = NULL,
    y = "Percent change in mean radiance"
  ) +
  theme_minimal()
```

```{r oahu-zones}
zone_daily <- extract_daily_means(r_all, oahu_zones, "zone") |>
  left_join(period_lookup, by = "date")

zone_summary <- zone_daily |>
  filter(period %in% c("Baseline", "Storm period")) |>
  group_by(area, period) |>
  summarize(mean_radiance = mean(radiance, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = period, values_from = mean_radiance) |>
  mutate(
    absolute_change = `Storm period` - Baseline,
    percent_change = 100 * absolute_change / Baseline
  )

zone_summary |>
  gt() |>
  fmt_number(columns = c(Baseline, `Storm period`, absolute_change), decimals = 2) |>
  fmt_number(columns = percent_change, decimals = 1) |>
  cols_label(area = "Oahu zone", `Storm period` = "Storm-period mean") |>
  tab_header(title = "Oahu Sub-area Baseline Versus Storm-Period Radiance")
```

```{r darkening-hotspots}
darkening_class <- classify(
  z_score_meaningful,
  rcl = matrix(
    c(
      -Inf, -2, 1,
      -2, -1, 2,
      -1, 1, 3,
      1, 2, 4,
      2, Inf, 5
    ),
    ncol = 3,
    byrow = TRUE
  )
)

levels(darkening_class) <- data.frame(
  value = 1:5,
  category = c(
    "Unusually darker",
    "Moderately darker",
    "Near baseline",
    "Moderately brighter",
    "Unusually brighter"
  )
)

ggplot() +
  geom_spatraster(data = darkening_class) +
  geom_sf(
    data = oahu_zones,
    fill = NA,
    color = "black",
    linewidth = 0.3
  ) +
  scale_fill_manual(
    values = c(
      "Unusually darker" = "#8B0000",
      "Moderately darker" = "#E76F51",
      "Near baseline" = "#F1F1F1",
      "Moderately brighter" = "#8EC07C",
      "Unusually brighter" = "#1B7837"
    ),
    na.value = "grey90",
    name = NULL
  ) +
  labs(
    title = "Oahu: Nighttime-Light Darkening Hotspots",
    subtitle = "Classification uses storm-period change relative to each pixel's pre-storm variability"
  ) +
  coord_sf() +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```
